---
title: "nestDevelopment"
output: pdf_document
---

### load packages
```{r}
library(tidyverse)
library(ggplot2)
library(car)
library(ggpubr)
library(MuMIn)
library(MASS)
library(plyr)
library(lme4)
```

Nests were originally coded as QA (i.e. Queen Alone), QT (i.e. Queen and Three workers), and QF (i.e. Queen and Five workers). group codes were changed to subsocial, eusocial (3W), and eusocial (5W), respectively, for the manuscript.

## load data
```{r}
# wd must be set to where this doc lives (chapter1/analyses)

# nest data
nest_data <- read.csv("../data/NestDissections.csv", header = TRUE)

# wing length data
wing_data <- read.csv("../data/MaleWingMeasurements.csv", header = TRUE)

# master data with natal colonies
master_data <- read.csv("../data/cleanMasterData.csv")
```

## clean and reformat data
```{r}
# filter data to only include nests from cleanMasterData (filtering was already done in cleanData file)
nest_data <- nest_data %>% subset(nestID %in% master_data$nestID)
wing_data <- wing_data %>% subset(nestID %in% master_data$nestID)

# merge nest_data with master_data 
nest_data <- join(nest_data, master_data, by = "nestID")

# add bee identifier to wing_data
wing_data$bee <- 1:nrow(wing_data)

# add nestTYPE column to all datasets, add columns for number of workers, number of queens, and number of bees
qChar <- sapply(strsplit(nest_data$nestID, ""), function(x) {x[1]} )
wChar <- sapply(strsplit(nest_data$nestID, ""), function(x) {x[2]} )
nest_data$nestTYPE <- paste0(qChar, wChar)
nest_data$nWorkers <- wChar
nest_data$nWorkers[nest_data$nWorkers == "F"] <- 5
nest_data$nWorkers[nest_data$nWorkers == "T"] <- 3
nest_data$nWorkers[nest_data$nWorkers == "A"] <- 0
nest_data$nWorkers <- as.numeric(nest_data$nWorkers)
nest_data$nBees <- nest_data$nWorkers + 1
nest_data

qChar <- sapply(strsplit(wing_data$nestID, ""), function(x) {x[1]} )
wChar <- sapply(strsplit(wing_data$nestID, ""), function(x) {x[2]} )
wing_data$nestTYPE <- paste0(qChar, wChar)
wing_data$nWorkers <- wChar
wing_data$nWorkers[wing_data$nWorkers == "F"] <- 5
wing_data$nWorkers[wing_data$nWorkers == "T"] <- 3
wing_data$nWorkers[wing_data$nWorkers == "A"] <- 0
wing_data$nWorkers <- as.numeric(wing_data$nWorkers)
wing_data$nBees <- wing_data$nWorkers + 1
wing_data

# add queen natal colony to wing_data
wing_data$Qnatal <- NA
for (i in 1:nrow(wing_data)) {
  for (j in 1:nrow(nest_data)) {
    if (wing_data$nestID[i] == nest_data$nestID[j]) {
      wing_data$Qnatal[i] <- nest_data$Qnatal[j]
    }
  }
}

# sum total number of brood in nest_data
nest_data$blackLarvae <- as.numeric(nest_data$blackLarvae)
nest_data$males <- as.numeric(nest_data$males)
nest_data$nBrood <- rowSums(nest_data[, c("e1", "e2", "e3", "e4", "e5", "e6", "e7", "e8", "e9", "e10", "e11", "e12", 
                                          "larvae", "blackLarvae", "pupae", "prePupae", "males")], na.rm = TRUE)

# sum total number of larvae in nest_data
nest_data$totLarvae <- rowSums(nest_data[, c("larvae", "blackLarvae", "prePupae")], na.rm = TRUE)

# sum total number of eggs in nest_data
nest_data$eggs <- rowSums(nest_data[, c("e1", "e2", "e3", "e4", "e5", "e6", "e7", "e8", "e9", "e10", "e11", "e12")], na.rm = TRUE)
```

# mixed models

### offspring size: p < 0.001
data is mostly normal
```{r}
# visualize the distribution
ggdensity(wing_data$beeAvg)
# normal 
qqp(wing_data$beeAvg, "norm")

# define model
maleSize <- lmer(beeAvg ~ nestTYPE + (1|Qnatal), data = wing_data)

# check for significance
tab_model(maleSize)
posthoc1 <- glht(maleSize, linfct = mcp(nestTYPE = "Tukey"))
summary(posthoc1)

# PLOT

# summarize mean, standard deviation and standard error of the mean
plotData <- wing_data
plotData$nWorkers <- factor(plotData$nWorkers, levels = c("0", "3", "5"))
plotData <- plotData[!is.na(plotData$beeAvg), ]
plotData
summ <- ddply(plotData, c("nWorkers"), summarise,
      mean = mean(beeAvg), sd = sd(beeAvg),
      sem = sd(beeAvg)/sqrt(length(beeAvg)))
summ

## make barplot with error bars
ggplot() +
  geom_bar(data = summ, aes(x = nWorkers, y = mean), stat = "identity", width = 0.5) +
  geom_errorbar(data = summ, aes(x = nWorkers, y = mean, ymin = mean - sem, ymax = mean + sem), width = 0.1) +
  theme(text = element_text(size=20)) +
  theme(legend.position="none") +
  labs(x = "Number of workers", 
       y = "Male wing cell length (mm)") +
  ggsave("maleSize.jpeg", path = "../figures")
```

### # brood total: p < 0.001
data is mostly normal
```{r}
# visualize the distribution
ggdensity(nest_data$nBrood)
# normal 
qqp(nest_data$nBrood, "norm")

# define model
nBrood <- lmer(nBrood ~ nestTYPE + (1|Qnatal), data = nest_data)

# check for significance
tab_model(nBrood)

posthoc1 <- glht(nBrood, linfct = mcp(nestTYPE = "Tukey"))
summary(posthoc1)

# PLOT

# summarize mean, standard deviation and standard error of the meand
plotData <- nest_data
plotData$nWorkers <- factor(plotData$nWorkers, levels = c("0", "3", "5"))
plotData <- plotData[!is.na(plotData$males), ]
summ <- ddply(plotData, c("nWorkers"), summarise,
      mean = mean(nBrood), sd = sd(nBrood),
      sem = sd(nBrood)/sqrt(length(nBrood)))
summ

## make barplot with error bars
ggplot() +
  geom_bar(data = summ, aes(x = nWorkers, y = mean), stat = "identity", width = 0.5) +
  geom_errorbar(data = summ, aes(x = nWorkers, y = mean, ymin = mean - sem, ymax = mean + sem), width = 0.1) +
  theme(text = element_text(size=20)) +
  theme(legend.position="none") +
  labs(x = "Number of workers", 
       y = "Number of brood items") +
  ggsave("nBrood.jpeg", path = "../figures")
```

### # eclosed adults: NS
data is normalish? not quite, though
```{r}
# visualize the distribution
ggdensity(nest_data$males)
# normal 
qqp(nest_data$males, "norm")

# define model
nMales <- lmer(males ~ nestTYPE + (1|Qnatal), data = nest_data)

# check for significance
tab_model(nMales)

posthoc1 <- glht(nMales, linfct = mcp(nestTYPE = "Tukey"))
summary(posthoc1)

# PLOT

# summarize mean, standard deviation and standard error of the meand
plotData <- nest_data
plotData$nWorkers <- factor(plotData$nWorkers, levels = c("0", "3", "5"))
plotData <- plotData[!is.na(plotData$males),]
summ <- ddply(plotData, c("nWorkers"), summarise,
      mean = mean(males), sd = sd(males),
      sem = sd(males)/sqrt(length(males)))
summ

## make barplot with error bars
ggplot() +
  geom_bar(data = summ, aes(x = nWorkers, y = mean), stat = "identity", width = 0.5) +
  geom_errorbar(data = summ, aes(x = nWorkers, y = mean, ymin = mean - sem, ymax = mean + sem), width = 0.1) +
  theme(text = element_text(size=20)) +
  theme(legend.position="none") +
  labs(x = "Number of workers", 
       y = "Number of males") +
  ggsave("nMales.jpeg", path = "../figures")
```


### # pupae: p < 0.001
data is mostly normal
```{r}
# visualize the distribution
ggdensity(nest_data$pupae)
# normal 
qqp(nest_data$pupae, "norm")

# define model
nPupae <- lmer(pupae ~ nestTYPE + (1|Qnatal), data = nest_data)

# check for significance
tab_model(nPupae)

posthoc1 <- glht(nPupae, linfct = mcp(nestTYPE = "Tukey"))
summary(posthoc1)

# PLOT

# summarize mean, standard deviation and standard error of the meand
plotData <- nest_data
plotData$nWorkers <- factor(plotData$nWorkers, levels = c("0", "3", "5"))
plotData <- plotData[!is.na(plotData$pupae),]
plotData
summ <- ddply(plotData, c("nWorkers"), summarise,
      mean = mean(pupae), sd = sd(pupae),
      sem = sd(pupae)/sqrt(length(pupae)))
summ

## make barplot with error bars
ggplot() +
  geom_bar(data = summ, aes(x = nWorkers, y = mean), stat = "identity", width = 0.5) +
  geom_errorbar(data = summ, aes(x = nWorkers, y = mean, ymin = mean - sem, ymax = mean + sem), width = 0.1) +
  theme(text = element_text(size=20)) +
  theme(legend.position="none") +
  labs(x = "Number of workers", 
       y = "Number of pupae") +
  ggsave("nPupae.jpeg", path = "../figures")
```

### # larvae + prepupae: NS
data is mostly normal
```{r}
# visualize the distribution
ggdensity(nest_data$totLarvae)
# normal 
qqp(nest_data$totLarvae, "norm")

# define model
totLarv <- lmer(totLarvae ~ nestTYPE + (1|Qnatal), data = nest_data)

# check for significance
tab_model(totLarv)

posthoc1 <- glht(totLarv, linfct = mcp(nestTYPE = "Tukey"))
summary(posthoc1)

# PLOT

# summarize mean, standard deviation and standard error of the meand
plotData <- nest_data
plotData$nWorkers <- factor(plotData$nWorkers, levels = c("0", "3", "5"))
plotData <- plotData[!is.na(plotData$totLarvae), ]
plotData
summ <- ddply(plotData, c("nWorkers"), summarise,
      mean = mean(totLarvae), sd = sd(totLarvae),
      sem = sd(totLarvae)/sqrt(length(totLarvae)))
summ

## make barplot with error bars
ggplot() +
  geom_bar(data = summ, aes(x = nWorkers, y = mean), stat = "identity", width = 0.5) +
  geom_errorbar(data = summ, aes(x = nWorkers, y = mean, ymin = mean - sem, ymax = mean + sem), width = 0.1) +
  theme(text = element_text(size=20)) +
  theme(legend.position="none") +
  labs(x = "Number of workers", 
       y = "Number of larvae") +
  ggsave("nTotLarvae.jpeg", path = "../figures")
```

### # eggs: NS
data is mostly normal
```{r}
# visualize the distribution
ggdensity(nest_data$eggs)
# normal 
qqp(nest_data$eggs, "norm")

# define model
egg <- lmer(eggs ~ nestTYPE + (1|Qnatal), data = nest_data)

# check for significance
tab_model(egg)

# PLOT

# summarize mean, standard deviation and standard error of the meand
plotData <- nest_data
plotData$nWorkers <- factor(plotData$nWorkers, levels = c("0", "3", "5"))
plotData <- plotData[!is.na(plotData$eggs), ]
plotData
summ <- ddply(plotData, c("nWorkers"), summarise,
      mean = mean(eggs), sd = sd(eggs),
      sem = sd(eggs)/sqrt(length(eggs)))
summ

## make barplot with error bars
ggplot() +
  geom_bar(data = summ, aes(x = nWorkers, y = mean), stat = "identity", width = 0.5) +
  geom_errorbar(data = summ, aes(x = nWorkers, y = mean, ymin = mean - sem, ymax = mean + sem), width = 0.1) +
  theme(text = element_text(size=20)) +
  theme(legend.position="none") +
  labs(x = "Number of workers", 
       y = "Number of eggs") +
  ggsave("nEggs.jpeg", path = "../figures")
```

<!-- ### number of days to lay first eggs: p < 0.001 -->
<!-- data is approximately normal -->
<!-- ```{r} -->
<!-- # visualize the distribution -->
<!-- ggdensity(dev_data$DaysToEggs) -->
<!-- # normal  -->
<!-- qqp(dev_data$DaysToEggs, "norm") -->
<!-- # # lnorm means lognormal -->
<!-- # qqp(dev_data$DaysToEggs, "lnorm") -->
<!-- # # gamma must be positive numbers -->
<!-- # gamma <- fitdistr(dev_data$DaysToEggs, "gamma") -->
<!-- # qqp(dev_data$DaysToEggs, "gamma", shape = gamma$estimate[[1]], rate = gamma$estimate[[2]]) -->

<!-- # define model -->
<!-- dayEggs <- lm(DaysToEggs ~ nWorkers, data = dev_data) -->

<!-- # check for significance -->
<!-- anova(dayEggs) -->

<!-- # PLOT -->

<!-- # summarize mean, standard deviation and standard error of the meand -->
<!-- plotData <- dev_data -->
<!-- plotData$nWorkers <- factor(plotData$nWorkers, levels = c("0", "3", "5")) -->
<!-- plotData <- plotData[!is.na(plotData$DaysToEggs), ] -->
<!-- plotData -->
<!-- summ <- ddply(plotData, c("nWorkers"), summarise, -->
<!--       mean = mean(DaysToEggs), sd = sd(DaysToEggs), -->
<!--       sem = sd(DaysToEggs)/sqrt(length(DaysToEggs))) -->
<!-- summ -->

<!-- ## make barplot with error bars -->
<!-- ggplot() + -->
<!--   geom_bar(data = summ, aes(x = nWorkers, y = mean), stat = "identity", width = 0.5) + -->
<!--   geom_errorbar(data = summ, aes(x = nWorkers, y = mean, ymin = mean - sem, ymax = mean + sem), width = 0.1) + -->
<!--   theme(text = element_text(size=20)) + -->
<!--   theme(legend.position="none") + -->
<!--   labs(x = "Number of workers",  -->
<!--        y = "Number of days until first eggs observed") + -->
<!--   ggsave("daysToEggs.jpeg", path = "../figures") -->
<!-- ``` -->

### # days from eggs to eclosion: NS 
data is approximately normal
```{r}
# visualize the distribution
ggdensity(dev_data$EggsToEclosion)
# normal 
qqp(dev_data$EggsToEclosion, "norm")
# # lnorm means lognormal
# qqp(dev_data$EggsToEclosion, "lnorm")
# # gamma must be positive numbers
# gamma <- fitdistr(dev_data$EggsToEclosion, "gamma")
# qqp(dev_data$EggsToEclosion, "gamma", shape = gamma$estimate[[1]], rate = gamma$estimate[[2]])

# define model
dayEclos <- lmer(EggsToEclosion ~ nestTYPE + (1|Qnatal), data = dev_data)

# check for significance
tab_model(dayEclos)

# PLOT

# summarize mean, standard deviation and standard error of the meand
plotData <- dev_data
plotData$nWorkers <- factor(plotData$nWorkers, levels = c("0", "3", "5"))
plotData <- plotData[!is.na(plotData$EggsToEclosion), ]
plotData
summ <- ddply(plotData, c("nWorkers"), summarise,
      mean = mean(EggsToEclosion), sd = sd(EggsToEclosion),
      sem = sd(EggsToEclosion)/sqrt(length(EggsToEclosion)))
summ

## make barplot with error bars
ggplot() +
  geom_bar(data = summ, aes(x = nWorkers, y = mean), stat = "identity", width = 0.5) +
  geom_errorbar(data = summ, aes(x = nWorkers, y = mean, ymin = mean - sem, ymax = mean + sem), width = 0.1) +
  theme(text = element_text(size=20)) +
  theme(legend.position="none") +
  labs(x = "Number of workers", 
       y = "Number of days from eggs to eclosion") +
  ggsave("eggsToEclosion.jpeg", path = "../figures")
```
